---
title: "OC SurgeWatch: Bayesian Surge Capacity Analysis for Orange County Hospital Disaster Preparedness"
author: "Joe T. Nguyen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

# Libraries

```{r libraries}
library(tidyverse)
library(brms)
library(tidybayes)
library(ggdist)
library(janitor)
library(readxl)
library(loo)
library(bayesplot)
library(posterior)
```

---

# Load Data & Identify Schema Shifts

```{r schema_shifts}
files <- list.files(path = "data", pattern = "pddpivot\\.csv$", full.names = TRUE)

# Snapshot column names by year for key modeling variables
schema_data <- map_df(files, function(path) {
  year_val <- as.numeric(str_extract(basename(path), "\\d{4}"))
  cols <- names(read_csv(path, n_max = 0, show_col_types = FALSE))
  tibble(year = year_val, variable = cols, present = 1) %>%
    filter(str_detect(variable, regex(
      "^(county|oshpd|age_6|age_7|age_8|toc_acute|payer_med|control|licensed)",
      ignore_case = TRUE
    )))
})

# Variables present in some years but not all
shifters <- schema_data %>%
  group_by(variable) %>%
  mutate(year_count = n_distinct(year)) %>%
  filter(year_count < length(files)) %>%
  ungroup()

ggplot(shifters, aes(x = factor(year), y = variable)) +
  geom_tile(aes(fill = factor(present)), color = "white") +
  scale_fill_manual(values = c("1" = "#e34a33"), guide = "none") +
  theme_minimal() +
  labs(
    title = "HCAI PDD Variable Name Shifts (2018–2024)",
    subtitle = "Key variables only — red = present that year",
    x = "Reporting Year", y = "Variable Name"
  ) +
  theme(
    axis.text.y      = element_text(size = 9),
    panel.grid.major = element_blank(),
    plot.title       = element_text(face = "bold")
  )
```

**Schema notes:**

Primary modeling variables (`toc_acute`, `age_60_69`, `age_70_79`, `age_80_`, `licensed_bed_size`, `OSHPD_ID`, `CONTROL_TYPE_DESC`) are consistently named across all 7 years. Two name shifts require bridging:

- `COUNTY_NAME` → `COUNTY` starting 2021
- `Payer_Medi_Cal` → `Payer_MediCal` in 2024

Case changes (`LICENSED_BED_SIZE` → `licensed_bed_size`) normalize automatically via `tolower()`.

---

# Preprocess Data

```{r preprocess}
# Column name bridges applied after tolower()
hcai_bridge <- c(
  "county"         = "county_name",  # pre-2021 name
  "payer_medi_cal" = "payer_medical" # 2024 name
)

load_pivot_year <- function(path) {
  df <- read_csv(path, col_types = cols(.default = "c"), show_col_types = FALSE)
  names(df) <- tolower(names(df))
  df <- df %>% rename(any_of(tolower(hcai_bridge)))
  df$data_year <- as.integer(str_extract(basename(path), "\\d{4}"))
  df %>%
    select(
      data_year, oshpd_id, facility_name, county,
      control_type_desc, licensed_bed_size,
      any_of(c("toc_acute", "age_60_69", "age_70_79", "age_80_",
               "payer_medicare", "adm_emergency", "adm_elective", "dsch_days"))
    )
}

# Excluded facility types (outside scope of adult acute-care surge analysis):
#   Pediatric: CHOC, Children's at Mission, Healthbridge
#   Rehab/LTACH: Kindred Brea, Kindred Westminster, Encompass Health Tustin
#   Psychiatric/substance abuse: College Hospital CM, Newport Bay, Laguna Treatment, Aliso Ridge
#   Elective specialty: Hoag Orthopedic Institute
#   Not a designated ERC: Chapman Global Medical Center
#   Incomplete panel: Anaheim Community Hospital (only 2022–2024)
exclude_patterns <- regex(
  paste(
    "children", "pediatric", "rehabilitation", "rehab",
    "psychiatric", "behavioral",
    "newport bay", "college hospital costa mesa", "laguna treatment",
    "kindred", "orthopedic", "chapman global", "anaheim community hospital",
    sep = "|"
  ),
  ignore_case = TRUE
)

pdd_raw <- map(files, load_pivot_year) %>%
  bind_rows() %>%
  filter(str_detect(toupper(county), "ORANGE")) %>%
  filter(!str_detect(facility_name, exclude_patterns)) %>%
  mutate(across(c(age_60_69, age_70_79, age_80_, toc_acute,
                  payer_medicare, adm_emergency, adm_elective, dsch_days),
                ~ as.numeric(.x)))

# Panel coverage audit — grouped by oshpd_id to handle mid-panel renames
panel_coverage <- pdd_raw %>%
  group_by(oshpd_id) %>%
  summarise(
    facility_name = last(facility_name[order(data_year)]),
    years_present = n_distinct(data_year),
    years         = paste(sort(unique(data_year)), collapse = ", "),
    toc_missing   = sum(is.na(toc_acute) | toc_acute == 0),
    age_missing   = sum(is.na(age_60_69) | is.na(age_70_79) | is.na(age_80_)),
    .groups = "drop"
  ) %>%
  arrange(years_present, oshpd_id)

cat("=== Panel Coverage Summary ===\n")
print(panel_coverage, n = Inf)

cat("\nFacilities by year count:\n")
print(table(panel_coverage$years_present))
```

---

# Derive Modelling Variables

```{r derive_variables}
pdd_panel <- pdd_raw %>%
  mutate(
    # Time index centered at 2018 (range 0–6)
    year_idx = data_year - 2018L,

    # Elderly admissions burden (60+)
    y_60plus     = age_60_69 + age_70_79 + age_80_,
    share_60plus = y_60plus / toc_acute,

    # Bed capacity midpoint — maps licensed_bed_size category to approximate count
    bed_midpoint = case_when(
      licensed_bed_size == "50-99"   ~ 75,
      licensed_bed_size == "100-149" ~ 125,
      licensed_bed_size == "150-199" ~ 175,
      licensed_bed_size == "200-299" ~ 250,
      licensed_bed_size == "300-499" ~ 400,
      TRUE ~ NA_real_
    ),

    # Occupancy rate; capped at 1.0 (values >1 reflect bed midpoint undercounting)
    occupancy = pmin(dsch_days / (bed_midpoint * 365), 1, na.rm = FALSE),

    # Ownership — EDA only; random effects absorb mid-panel changes
    ownership = case_when(
      str_detect(control_type_desc, regex("investor", ignore_case = TRUE))                              ~ "ForProfit",
      str_detect(control_type_desc, regex("non-profit|church", ignore_case = TRUE))                     ~ "Nonprofit",
      str_detect(control_type_desc, regex("university|state|city|county|district", ignore_case = TRUE)) ~ "Government",
      TRUE ~ "Other"
    ),
    ownership = factor(ownership, levels = c("Nonprofit", "ForProfit", "Government")),

    emerg_share    = adm_emergency / toc_acute,
    elective_share = adm_elective  / toc_acute,
    medicare_share = payer_medicare / toc_acute
  )

# Report any rows with missing/zero primary outcome
pdd_panel %>%
  filter(is.na(toc_acute) | toc_acute == 0) %>%
  select(data_year, oshpd_id, facility_name, toc_acute) %>%
  { if (nrow(.) > 0) { cat("Rows dropped (missing/zero toc_acute):\n"); print(.) }
    else cat("No rows dropped — toc_acute complete.\n") }

pdd_panel <- pdd_panel %>%
  filter(!is.na(toc_acute), toc_acute > 0)

cat("\n", nrow(pdd_panel), "rows |",
    n_distinct(pdd_panel$oshpd_id), "facilities\n")

glimpse(pdd_panel)
```

---

# Data Validation

```{r data_validation_model_vars}
# Missing value heatmap — core modeling variables
m_vars <- c("toc_acute", "year_idx", "oshpd_id")

pdd_panel %>%
  group_by(data_year) %>%
  summarise(across(all_of(m_vars), ~ mean(is.na(.)))) %>%
  pivot_longer(-data_year, names_to = "variable", values_to = "pct_missing") %>%
  mutate(variable = factor(variable, levels = m_vars)) %>%
  ggplot(aes(x = factor(data_year), y = variable, fill = pct_missing)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "#ef3b2c", labels = scales::percent_format()) +
  theme_minimal() +
  labs(
    title = "Missing Value Density — Core Variables",
    subtitle = "Primary outcome: toc_acute. All white = zero missing.",
    x = "Year", y = NULL, fill = "% Missing"
  )
```

```{r data_validation_scorecard}
# Missing value heatmap — scorecard features
scorecard_vars <- c("y_60plus", "share_60plus", "occupancy",
                    "emerg_share", "elective_share", "medicare_share")

pdd_panel %>%
  group_by(data_year) %>%
  summarise(across(all_of(scorecard_vars), ~ mean(is.na(.)))) %>%
  pivot_longer(-data_year, names_to = "variable", values_to = "pct_missing") %>%
  mutate(variable = factor(variable, levels = scorecard_vars)) %>%
  ggplot(aes(x = factor(data_year), y = variable, fill = pct_missing)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "#ef3b2c", labels = scales::percent_format()) +
  theme_minimal() +
  labs(
    title = "Missing Value Density — Scorecard Feature Variables",
    subtitle = "occupancy is NA where dsch_days is missing.",
    x = "Year", y = NULL, fill = "% Missing"
  )
```

```{r data_validation_raw_indicators}
# Missing value heatmap — raw surge and financial indicators
raw_indicators <- c("payer_medicare", "adm_emergency", "adm_elective", "dsch_days")

pdd_panel %>%
  group_by(data_year) %>%
  summarise(across(all_of(raw_indicators), ~ mean(is.na(.)))) %>%
  pivot_longer(-data_year, names_to = "variable", values_to = "pct_missing") %>%
  ggplot(aes(x = factor(data_year), y = variable, fill = pct_missing)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "#ef3b2c", labels = scales::percent_format()) +
  theme_minimal() +
  labs(
    title = "Missing Value Density — Raw Surge & Financial Indicators",
    subtitle = "adm_emergency/adm_elective missingness in 2020–2021 is MNAR (reporting waivers).",
    x = "Year", y = NULL, fill = "% Missing"
  )
```

---

# Save Processed Data

```{r save_processed}
dir.create("data/processed", showWarnings = FALSE)
saveRDS(pdd_panel, "data/processed/pdd_panel.rds")

cat("Saved: data/processed/pdd_panel.rds —",
    nrow(pdd_panel), "rows,",
    n_distinct(pdd_panel$oshpd_id), "facilities\n")
```
