---
title: "EDA"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

# Libraries

```{r libraries}
library(tidyverse)
library(ggdist)
library(knitr)
```

# Load Data

```{r load}
pdd <- readRDS("data/processed/pdd_panel.rds")

# Short display labels for plots
facility_labels <- c(
  "300225" = "Orange Coast Memorial",
  "301098" = "Anaheim Regional",
  "301175" = "UCI Fountain Valley",
  "301188" = "Anaheim Global",
  "301205" = "Hoag Newport Beach",
  "301209" = "Huntington Beach",
  "301234" = "La Palma",
  "301248" = "UCI Los Alamitos",
  "301258" = "South Coast Global",
  "301262" = "Mission Viejo",
  "301279" = "UCI Orange",
  "301283" = "Garden Grove",
  "301297" = "UCI Placentia Linda",
  "301317" = "Saddleback",
  "301337" = "Mission Laguna Beach",
  "301340" = "St. Joseph",
  "301342" = "St. Jude",
  "301357" = "Foothill Regional",
  "301379" = "West Anaheim",
  "301566" = "OC Global",
  "304409" = "Kaiser Anaheim"
)

pdd <- pdd %>%
  mutate(
    facility_label = facility_labels[oshpd_id],
    period = case_when(
      data_year %in% c(2018, 2019) ~ "Pre-COVID",
      data_year %in% c(2020, 2021) ~ "COVID",
      data_year >= 2022            ~ "Post-COVID"
    ),
    period = factor(period, levels = c("Pre-COVID", "COVID", "Post-COVID"))
  )

# Reusable COVID shading band for time-series plots
covid_band <- annotate("rect", xmin = 2019.5, xmax = 2021.5,
                       ymin = -Inf, ymax = Inf,
                       fill = "#d7191c", alpha = 0.08)

# Base theme with bold titles
theme_oc <- function(...) {
  theme_minimal() +
    theme(
      plot.title    = element_text(face = "bold"),
      plot.subtitle = element_text(lineheight = 1.3)
    ) +
    theme(...)
}
```

------------------------------------------------------------------------

# 1. Outcome Distribution

```{r outcome_histogram}
ggplot(pdd, aes(x = toc_acute)) +
  geom_histogram(bins = 30, fill = "#2c7bb6", color = "white", linewidth = 0.3) +
  scale_x_continuous(labels = scales::comma) +
  theme_oc() +
  labs(
    title = "Distribution of Annual Acute Admissions (toc_acute)",
    subtitle = "Bimodal: small community hospitals cluster at 1k–5k; large regional centers at 10k–45k.\nA pooled intercept fits neither.",
    x = "Total Acute Admissions", y = "Count"
  )
```

```{r outcome_log}
ggplot(pdd, aes(x = toc_acute)) +
  geom_histogram(bins = 30, fill = "#2c7bb6", color = "white", linewidth = 0.3) +
  scale_x_log10(labels = scales::comma) +
  theme_oc() +
  labs(
    title = "Distribution of Annual Acute Admissions (toc_acute - log scale)",
    subtitle = "Bimodality on log scale — two facility sizes, not a single population.\n Random intercepts may be needed.",
    x = "Total Acute Admissions (log scale)", y = "Count"
  )
```

------------------------------------------------------------------------

# 2. Overdispersion

For a Poisson model, Var = Mean within each facility. We compare the observed within-facility variance to that expectation.

```{r overdispersion}
overdisp <- pdd %>%
  group_by(facility_label) %>%
  summarise(
    mean_toc    = mean(toc_acute),
    var_toc     = var(toc_acute),
    .groups = "drop"
  )

ggplot(overdisp, aes(x = mean_toc)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey50", linewidth = 0.8) +
  geom_point(aes(y = var_toc), color = "#d7191c", size = 3) +
  annotate("text", x = 34000, y = 3000000,
           label = "Poisson: Var = Mean", color = "grey40",
           size = 3.5, fontface = "italic") +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  theme_oc() +
  labs(
    title = "Variance vs. Mean by Facility Across Years (Overdispersion Check)",
    subtitle = "Poisson line is invisible at this scale — annual swings push observed variance way \n above the Poisson mean assumption. Negative Binomial is required.",
    x = "Within-Facility Mean (toc_acute)", y = "Within-Facility Variance"
  )
```

```{r overdispersion_table}
overdisp_wide <- pdd %>%
  group_by(facility_label) %>%
  summarise(
    `2018` = toc_acute[data_year == 2018],
    `2019` = toc_acute[data_year == 2019],
    `2020` = toc_acute[data_year == 2020],
    `2021` = toc_acute[data_year == 2021],
    `2022` = toc_acute[data_year == 2022],
    `2023` = toc_acute[data_year == 2023],
    `2024` = toc_acute[data_year == 2024],
    Mean     = round(mean(toc_acute)),
    Variance = round(var(toc_acute)),
    .groups = "drop"
  ) %>%
  arrange(Mean) %>%
  rename(Facility = facility_label)

overdisp_wide
```

------------------------------------------------------------------------

# 3. Between-Facility Scale

```{r facility_scale}
pdd %>%
  group_by(facility_label) %>%
  summarise(mean_toc = mean(toc_acute), .groups = "drop") %>%
  arrange(mean_toc) %>%
  mutate(facility_label = fct_inorder(facility_label)) %>%
  ggplot(aes(x = mean_toc, y = facility_label)) +
  geom_segment(aes(x = 0, xend = mean_toc, yend = facility_label),
               color = "grey80", linewidth = 0.6) +
  geom_point(color = "#2c7bb6", size = 3) +
  scale_x_continuous(labels = scales::comma) +
  theme_oc() +
  labs(
    title = "Mean Annual Acute Admissions by Facility (2018–2024)",
    subtitle = "From Mission Laguna Beach (~1,800) to Hoag Newport Beach (~40,000).\nThree natural size tiers visible — a pooled intercept would be meaningless.",
    x = "Mean Annual Admissions", y = "Facility"
  )
```

------------------------------------------------------------------------

# 4. Temporal Trends

```{r temporal_spaghetti}
sys_mean <- pdd %>%
  group_by(data_year) %>%
  summarise(mean_toc = mean(toc_acute), .groups = "drop")

ggplot(pdd, aes(x = data_year, y = toc_acute, group = facility_label)) +
  covid_band +
  geom_line(aes(color = ownership), alpha = 0.6, linewidth = 0.8) +
  geom_line(data = sys_mean, aes(y = mean_toc, group = 1, color = "System Mean"),
            linewidth = 1.2, linetype = "dashed") +
  scale_x_continuous(breaks = 2018:2024) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_manual(
    values = c(
      "Nonprofit"   = "#2c7bb6",
      "ForProfit"   = "#d7191c",
      "Government"  = "#1a9641",
      "System Mean" = "black"
    )
  ) +
  guides(color = guide_legend(
    override.aes = list(
      linetype  = c("solid", "solid", "solid", "dashed"),
      linewidth = c(0.8, 0.8, 0.8, 1.2),
      alpha     = c(0.6, 0.6, 0.6, 1)
    )
  )) +
  theme_oc() +
  labs(
    title = "Annual Admissions by Facility Over Time",
    subtitle = "Nonprofits (blue) dominate volume; for-profits (red) more COVID-sensitive.\n System mean dips during COVID years. Ownership changes exists.",
    x = "Year", y = "Total Annual Admissions", color = NULL
  )
```

```{r temporal_system}
pdd %>%
  group_by(data_year, period) %>%
  summarise(mean_toc = mean(toc_acute), sd_toc = sd(toc_acute), .groups = "drop") %>%
  ggplot(aes(x = data_year, y = mean_toc)) +
  covid_band +
  geom_ribbon(aes(ymin = mean_toc - sd_toc, ymax = mean_toc + sd_toc),
              fill = "#2c7bb6", alpha = 0.15) +
  geom_line(color = "#2c7bb6", linewidth = 1.2) +
  geom_point(aes(fill = period), shape = 21, size = 3.5, color = "white", stroke = 1) +
  scale_fill_manual(values = c(
    "Pre-COVID"  = "#1a9641",
    "COVID"      = "#d7191c",
    "Post-COVID" = "#2c7bb6"
  )) +
  scale_x_continuous(breaks = 2018:2024) +
  scale_y_continuous(labels = scales::comma) +
  theme_oc() +
  labs(
    title = "System-Wide Mean Acute Admissions by Year",
    subtitle = "System-level COVID dip was modest (~10%); full recovery by 2023.\nWide ribbon reflects 23× size heterogeneity across facilities — not forecast uncertainty.",
    x = "Year", y = "Mean Annual Acute Admissions", fill = "Period"
  )
```

------------------------------------------------------------------------

# 5. COVID Effect Heterogeneity

\% change from pre-COVID baseline (2018–2019) to COVID years (2020–2021), by facility.

```{r covid_effect}
pdd %>%
  filter(period != "Post-COVID") %>%
  group_by(facility_label, period) %>%
  summarise(mean_toc = mean(toc_acute), .groups = "drop") %>%
  pivot_wider(names_from = period, values_from = mean_toc) %>%
  mutate(
    pct_change = (COVID - `Pre-COVID`) / `Pre-COVID` * 100,
    direction  = if_else(pct_change >= 0, "Increase", "Decrease"),
    facility_label = fct_reorder(facility_label, pct_change)
  ) %>%
  ggplot(aes(x = pct_change, y = facility_label, fill = direction)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 0, color = "grey30", linewidth = 0.5) +
  scale_fill_manual(values = c("Decrease" = "#d7191c", "Increase" = "#1a9641")) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  theme_oc() +
  theme(legend.position = "none") +
  labs(
    title = "COVID Effect by Facility: % Change in Admissions\n(2020–21 vs 2018–19)",
    subtitle = "Only 3 of 21 facilities saw volume increase during COVID.\nRange: −42% to +24%",
    x = "% Change vs Pre-COVID Baseline", y = "Facility"
  )
```

------------------------------------------------------------------------

# 6. Scorecard Feature Distributions

Pre-COVID baseline averages (2018–2019) used throughout — less contaminated by COVID disruptions.

```{r scorecard_distributions}
scorecard_pre <- pdd %>%
  filter(period == "Pre-COVID") %>%
  group_by(facility_label) %>%
  summarise(
    `Elderly Share (60+)` = mean(share_60plus,   na.rm = TRUE),
    `Emergency Share`     = mean(emerg_share,    na.rm = TRUE),
    `Occupancy Rate`      = mean(occupancy,      na.rm = TRUE),
    `Medicare Share`      = mean(medicare_share, na.rm = TRUE),
    .groups = "drop"
  )

scorecard_pre %>%
  pivot_longer(-facility_label, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 10, fill = "#2c7bb6", color = "white", linewidth = 0.3) +
  facet_wrap(~ feature, scales = "free", nrow = 2) +
  theme_oc() +
  labs(
    title = "Scorecard Feature Distributions (Baseline, 2018–2019)",
    subtitle = "21 facilities. Each scorecard share is a proportion of total admissions;\n Reporting anomaly in Emergency Share exceeding total admissions.",
    x = "Proportion / Rate", y = "Facility Count"
  )
```

```{r scorecard_by_facility}
scorecard_pre %>%
  pivot_longer(-facility_label, names_to = "feature", values_to = "value") %>%
  mutate(facility_label = fct_reorder(facility_label, value, .fun = mean)) %>%
  ggplot(aes(x = value, y = facility_label)) +
  geom_point(color = "#2c7bb6", size = 2) +
  facet_wrap(~ feature, scales = "free_x", nrow = 1) +
  theme_oc(axis.text.y = element_text(size = 7)) +
  labs(
    title = "Scorecard Features by Facility (Pre-COVID Baseline)",
    subtitle = "Facilities sorted by mean feature value. Variation is substantial across \n all dimensions. Emergency Share > 1.0 likely reflect data/reporting anomalies. \n Occupancy rate > 1.0 likely due to licensed bed midpoint undercounting actual \n capacity.",
    x = "Proportion / Rate", y = "Facility"
  )
```

```{r scorecard_correlation}
cor_mat <- scorecard_pre %>%
  select(-facility_label) %>%
  cor(use = "complete.obs")

as.data.frame(as.table(cor_mat)) %>%
  rename(feature1 = Var1, feature2 = Var2, r = Freq) %>%
  ggplot(aes(x = feature1, y = feature2, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(r, 2)), size = 3.5) +
  scale_fill_gradient2(low = "#d7191c", mid = "white", high = "#2c7bb6",
                       midpoint = 0, limits = c(-1, 1)) +
  theme_oc(axis.text.x = element_text(angle = 20, hjust = 1)) +
  labs(
    title = "Scorecard Feature Correlation Matrix",
    subtitle = "Elderly Share and Medicare Share are nearly collinear (r = 0.92). \n Occupancy is moderately negative with all three (r ≈ −0.57).",
    x = NULL, y = NULL, fill = "r"
  )
```

---

# 7. Scorecard Data Quality Check

```{r scorecard_quality}
pdd %>%
  filter(emerg_share > 1 | occupancy > 1) %>%
  select(facility_label, data_year, emerg_share, occupancy,
         toc_acute, adm_emergency, dsch_days) %>%
  arrange(facility_label, data_year)
```

---

# 8. EDA Summary

- A clear bimodal distribution — small community hospitals (1k–5k admissions/year) and large regional centers (10k–45k) are essentially two different populations. This alone justifies the hierarchical model: a pooled intercept is meaningless, and partial pooling via random effects is the minimum viable approach.

- Overdispersion is severe — facility-level variance runs 250–1,000× the Poisson expectation, driven by the COVID shock compressing 7 years of temporal variation into a small panel. Negative Binomial is required.

- **The system-level COVID effect was modest (~10% dip).** Going in, we expected a dramatic system-wide collapse. Instead, the story is almost entirely about heterogeneity: 18 of 21 facilities declined, but 3 actually grew — Foothill Regional by +24%. The system mean barely moved because large facilities (which dominate the average) were relatively stable.

- **Elderly Share and Medicare Share are r = 0.92.** Expected correlation, not near-identity. These two variables are measuring essentially the same thing — older patients are disproportionately on Medicare.

- Q2 (how did COVID stress the system?) has a more nuanced answer than anticipated: the aggregate effect is small, but facility-level divergence is large and worth explaining. The action is in Q1 and Q3 — which facilities diverged most, and what predicted it.
