---
title: "Modeling"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Libraries

```{r libraries}
library(tidyverse)
library(brms)
library(tidybayes)
library(ggdist)
library(loo)
library(bayesplot)
library(posterior)

dir.create("models", showWarnings = FALSE)
```

------------------------------------------------------------------------

# Load Data

```{r load}
pdd <- readRDS("data/processed/pdd_panel.rds")

facility_labels <- c(
  "300225" = "Orange Coast Memorial",
  "301098" = "Anaheim Regional",
  "301175" = "UCI Fountain Valley",
  "301188" = "Anaheim Global",
  "301205" = "Hoag Newport Beach",
  "301209" = "Huntington Beach",
  "301234" = "La Palma",
  "301248" = "UCI Los Alamitos",
  "301258" = "South Coast Global",
  "301262" = "Mission Viejo",
  "301279" = "UCI Orange",
  "301283" = "Garden Grove",
  "301297" = "UCI Placentia Linda",
  "301317" = "Saddleback",
  "301337" = "Mission Laguna Beach",
  "301340" = "St. Joseph",
  "301342" = "St. Jude",
  "301357" = "Foothill Regional",
  "301379" = "West Anaheim",
  "301566" = "OC Global",
  "304409" = "Kaiser Anaheim"
)

pdd <- pdd %>%
  mutate(facility_label = facility_labels[oshpd_id])

cat(nrow(pdd), "rows |", n_distinct(pdd$oshpd_id), "facilities |",
    n_distinct(pdd$data_year), "years\n")
```

------------------------------------------------------------------------

# Prepare Scorecard Features

Baseline averages (2018-2019) for Elderly Share and Occupancy Rate. Z-scored so coefficients are comparable. Using the early baseline as the fixed reference point — in a real-time monitoring system, these would be updated periodically against the same baseline.

```{r scorecard_features}
scorecard <- pdd %>%
  filter(data_year %in% c(2018, 2019)) %>%
  group_by(oshpd_id) %>%
  summarise(
    share_60plus_raw = mean(share_60plus, na.rm = TRUE),
    occupancy_raw    = mean(occupancy,    na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    share_60plus_z = as.numeric(scale(share_60plus_raw)),
    occupancy_z    = as.numeric(scale(occupancy_raw))
  )

pdd <- pdd %>%
  left_join(scorecard %>% select(oshpd_id, share_60plus_z, occupancy_z),
            by = "oshpd_id")

cat("Scorecard features - baseline means and SDs:\n")
scorecard %>%
  summarise(
    across(ends_with("_raw"),
           list(mean = ~ round(mean(.), 3), sd = ~ round(sd(.), 3)))
  )
```

------------------------------------------------------------------------

# Train / Test Split

Training: 2018-2023 (6 years). Test: 2024 (1 year — held out entirely).

```{r train_test_split}
train <- pdd %>% filter(data_year <= 2023)
test  <- pdd %>% filter(data_year == 2024)

cat("Train:", nrow(train), "rows |", n_distinct(train$data_year), "years (",
    paste(sort(unique(train$data_year)), collapse = ", "), ")\n")
cat("Test: ", nrow(test),  "rows |", n_distinct(test$data_year),  "years (",
    paste(sort(unique(test$data_year)),  collapse = ", "), ")\n")
```

------------------------------------------------------------------------

# Priors

```{r priors}
# Q1 forecasting model (train only): base — no scorecard features
priors_base <- c(
  prior(normal(log(10000), 0.5), class = Intercept),
  prior(normal(0, 0.05),         class = b, coef = year_idx),
  prior(normal(0, 0.3),          class = sd),
  prior(gamma(4, 0.1),           class = shape)
)

# Q2 forecasting model (train only): scorecard features added
priors_scorecard <- c(
  prior(normal(log(10000), 0.5), class = Intercept),
  prior(normal(0, 0.05),         class = b, coef = year_idx),
  prior(normal(0, 0.5),          class = b, coef = share_60plus_z),
  prior(normal(0, 0.5),          class = b, coef = occupancy_z),
  prior(normal(0, 0.3),          class = sd),
  prior(gamma(4, 0.1),           class = shape)
)
```

------------------------------------------------------------------------

# Prior Predictive Check

```{r prior_check, cache=TRUE}
m_prior <- brm(
  toc_acute ~ year_idx + (1 | oshpd_id),
  data         = train,
  family       = negbinomial(),
  prior        = priors_base,
  sample_prior = "only",
  chains = 2, iter = 1000, warmup = 500,
  seed   = 42,
  file   = "models/m_prior_v2"
)

yrep <- posterior_predict(m_prior, ndraws = 50)
yrep_df <- as.data.frame(t(yrep)) %>%
  pivot_longer(everything(), names_to = "draw", values_to = "value")

ggplot() +
  geom_density(data = yrep_df, aes(x = value, group = draw),
               color = "#d7191c", alpha = 0.5, linewidth = 0.3) +
  geom_density(data = data.frame(value = train$toc_acute), aes(x = value),
               color = "black", linewidth = 1) +
  scale_x_continuous(labels = scales::comma) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(lineheight = 1.3)) +
  labs(title    = "Prior Predictive Check",
       subtitle = "Red = simulated (prior only); black = observed (train).\nSimulated outcomes should bracket the observed range without blowing up.",
       x = "Total Acute Admissions", y = "Density")
```

------------------------------------------------------------------------

# Q1: Facility Vulnerability Forecasting — Base Model

Fit on 2018-2023. Facility random intercepts capture baseline size differences. No scorecard features — this is the baseline predictive model.

```{r m_base, cache=TRUE}
m_base <- brm(
  toc_acute ~ year_idx + (1 | oshpd_id),
  data    = train,
  family  = negbinomial(),
  prior   = priors_base,
  chains  = 4, iter = 3000, warmup = 1500,
  control = list(adapt_delta = 0.95),
  seed    = 42,
  file    = "models/m_base_v3"
)

summary(m_base)
```

------------------------------------------------------------------------

# Q2: Facility Vulnerability Forecasting — Scorecard Model

Same structure as m_base but adds baseline occupancy rate and elderly share as fixed predictors. Tests whether structural facility characteristics improve our ability to forecast which facilities are under the most demand pressure.

```{r m_scorecard, cache=TRUE}
m_scorecard <- brm(
  toc_acute ~ year_idx + share_60plus_z + occupancy_z + (1 | oshpd_id),
  data    = train,
  family  = negbinomial(),
  prior   = priors_scorecard,
  chains  = 4, iter = 3000, warmup = 1500,
  control = list(adapt_delta = 0.95),
  seed    = 42,
  file    = "models/m_scorecard_v3"
)

summary(m_scorecard)

# Feature coefficients — practical interpretation
m_scorecard %>%
  spread_draws(b_share_60plus_z, b_occupancy_z) %>%
  pivot_longer(starts_with("b_"), names_to = "feature", values_to = "estimate") %>%
  mutate(
    feature    = recode(feature,
                        "b_share_60plus_z" = "Elderly Share (60+)",
                        "b_occupancy_z"    = "Occupancy Rate"),
    pct_change = (exp(estimate) - 1) * 100
  ) %>%
  group_by(feature) %>%
  median_qi(pct_change, .width = 0.90) %>%
  mutate(across(where(is.numeric), ~ round(., 1))) %>%
  transmute(
    Feature                   = feature,
    `Effect per 1-SD (median)` = paste0(pct_change, "%"),
    `90% Lower`               = paste0(.lower, "%"),
    `90% Upper`               = paste0(.upper, "%")
  )
```

```{r scorecard_coef_plot}
# Posterior distributions of scorecard feature effects (% change in admissions per 1-SD)
m_scorecard %>%
  spread_draws(b_share_60plus_z, b_occupancy_z) %>%
  pivot_longer(starts_with("b_"), names_to = "feature", values_to = "estimate") %>%
  mutate(
    feature    = recode(feature,
                        "b_share_60plus_z" = "Elderly Share (60+)",
                        "b_occupancy_z"    = "Occupancy Rate"),
    pct_change = (exp(estimate) - 1) * 100
  ) %>%
  ggplot(aes(x = pct_change, y = feature, fill = feature)) +
  stat_halfeye(.width = c(0.50, 0.90), point_interval = median_qi,
               alpha = 0.85, normalize = "panels") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey30", linewidth = 0.7) +
  scale_fill_manual(values = c("Elderly Share (60+)" = "#2c7bb6",
                               "Occupancy Rate"       = "#d7191c")) +
  theme_minimal() +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(lineheight = 1.3),
        legend.position = "none") +
  labs(
    title    = "Scorecard Feature Effects on Annual Admissions",
    subtitle = "Posterior distribution of % change in admissions per 1-SD increase in each feature.\nInner bar = 50% credible interval; outer bar = 90% credible interval. \nDashed line = no effect.",
    x = "% Change in Admissions per 1-SD", y = NULL
  )
```

------------------------------------------------------------------------

# Convergence Checks

```{r convergence}
models <- list(Base = m_base, Scorecard = m_scorecard)

map_df(names(models), function(nm) {
  draws     <- as_draws_df(models[[nm]])
  rhat_vals <- summarise_draws(draws, rhat)$rhat
  ess_vals  <- summarise_draws(draws, ess_bulk)$ess_bulk
  tibble(
    Model        = nm,
    max_rhat     = round(max(rhat_vals, na.rm = TRUE), 3),
    min_ess_bulk = round(min(ess_vals,  na.rm = TRUE))
  )
})
```

------------------------------------------------------------------------

# Posterior Predictive Evaluation on Test Set (2024)

Compare m_base and m_scorecard predictions against held-out actuals. Coverage: what % of actuals fall within the 90% posterior predictive interval. RMSE: root mean squared error on point estimates (posterior median).

```{r ppc_test}
pred_base      <- posterior_predict(m_base,      newdata = test)
pred_scorecard <- posterior_predict(m_scorecard, newdata = test)

eval_model <- function(pred_mat, label) {
  test %>%
    mutate(
      pred_median = apply(pred_mat, 2, median),
      pred_lower  = apply(pred_mat, 2, quantile, 0.05),
      pred_upper  = apply(pred_mat, 2, quantile, 0.95),
      residual    = toc_acute - pred_median,
      covered     = toc_acute >= pred_lower & toc_acute <= pred_upper,
      model       = label
    )
}

eval_base      <- eval_model(pred_base,      "Base")
eval_scorecard <- eval_model(pred_scorecard, "Scorecard")

# Summary metrics
bind_rows(eval_base, eval_scorecard) %>%
  group_by(model) %>%
  summarise(
    `90% PI Coverage` = scales::percent(mean(covered), accuracy = 0.1),
    RMSE              = round(sqrt(mean(residual^2))),
    MAE               = round(mean(abs(residual))),
    .groups = "drop"
  )
```

```{r ppc_test_plot}
# Predicted vs actual by facility
bind_rows(eval_base, eval_scorecard) %>%
  mutate(data_year = factor(data_year)) %>%
  ggplot(aes(x = pred_median, y = toc_acute, color = model)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_manual(values = c("Base" = "#2c7bb6", "Scorecard" = "#d7191c")) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(lineheight = 1.3)) +
  labs(
    title    = "Predicted vs Actual Acute Admissions — Test Set (2024)",
    subtitle = "Points on the dashed line = perfect prediction.\nScatter above/below = over/under-prediction per facility-year.",
    x = "Predicted (posterior median)", y = "Actual", color = "Model"
  )
```

```{r ppc_interval_plot}
# Per-facility 90% PI plot — shows coverage and direction of error
eval_scorecard %>%
  mutate(
    facility_label = fct_reorder(facility_label, pred_median),
    covered        = toc_acute >= pred_lower & toc_acute <= pred_upper
  ) %>%
  ggplot(aes(y = facility_label)) +
  geom_segment(aes(x = pred_lower, xend = pred_upper, yend = facility_label),
               color = "grey70", linewidth = 1) +
  geom_point(aes(x = pred_median), color = "#d7191c", size = 2.5) +
  geom_point(aes(x = toc_acute, shape = covered), size = 2.5, color = "black") +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 4),
                     labels = c("TRUE" = "Inside PI", "FALSE" = "Outside PI")) +
  scale_x_continuous(labels = scales::comma) +
  theme_minimal() +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(lineheight = 1.3),
        legend.position = "bottom") +
  labs(
    title    = "Posterior Credible Intervals vs Actuals — Test Set (2024)",
    subtitle = "Red dot = predicted median; black dot = actual 2024 admissions.\nFilled circle = actual falls inside 90% CI; X = outside. \nScorecard model, all 21 facilities.",
    x = "Annual Admissions", y = NULL, shape = "Actual"
  )
```

```{r ppc_residuals_plot}
# Residuals by facility — signed error (actual - predicted)
eval_scorecard %>%
  mutate(
    facility_label = fct_reorder(facility_label, residual),
    direction      = if_else(residual >= 0, "Under-predicted", "Over-predicted")
  ) %>%
  ggplot(aes(x = residual, y = facility_label, fill = direction)) +
  geom_col(width = 0.6) +
  geom_vline(xintercept = 0, linewidth = 0.8, color = "grey30") +
  scale_x_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("Under-predicted" = "#d7191c", "Over-predicted" = "#2c7bb6")) +
  theme_minimal() +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(lineheight = 1.3),
        legend.position = "bottom") +
  labs(
    title    = "Forecast Residuals by Facility — Test Set (2024)",
    subtitle = "Residual = actual - predicted median. Red = model undershot actual volume; blue = model overshot.\nLarge positive residuals flag facilities the model systematically underestimates.",
    x = "Residual (Actual - Predicted)", y = NULL, fill = NULL
  )
```

------------------------------------------------------------------------

# Q2: Feature Contribution to Predictive Accuracy

LOO comparison of m_base vs m_scorecard on the training data. Positive elpd_diff for Scorecard = scorecard features improve prediction.

```{r feature_contribution}
loo_base      <- loo(m_base)
loo_scorecard <- loo(m_scorecard)

loo_compare(loo_base, loo_scorecard)
```

------------------------------------------------------------------------

# Q1: Facility Vulnerability Rankings (2024)

Facilities ranked by predicted 2024 acute admissions volume (posterior median). Higher predicted volume + higher baseline occupancy = greater surge exposure. The 90% posterior predictive interval reflects forecast uncertainty per facility.

```{r vulnerability_rankings}
vulnerability <- eval_scorecard %>%
  group_by(facility_label) %>%
  summarise(
    pred_median   = mean(pred_median),
    pred_lower    = mean(pred_lower),
    pred_upper    = mean(pred_upper),
    actual_mean   = mean(toc_acute),
    residual_mean = mean(residual),
    .groups = "drop"
  ) %>%
  left_join(
    scorecard %>%
      mutate(facility_label = facility_labels[as.character(oshpd_id)]) %>%
      select(facility_label, occupancy_raw, share_60plus_raw),
    by = "facility_label"
  ) %>%
  mutate(
    occupancy_tier = case_when(
      occupancy_raw >= quantile(occupancy_raw, 0.67, na.rm = TRUE) ~ "High",
      occupancy_raw >= quantile(occupancy_raw, 0.33, na.rm = TRUE) ~ "Medium",
      TRUE ~ "Low"
    ),
    occupancy_tier = factor(occupancy_tier, levels = c("High", "Medium", "Low")),
    facility_label = fct_reorder(facility_label, pred_median)
  ) %>%
  arrange(desc(pred_median))

ggplot(vulnerability, aes(x = pred_median, y = facility_label)) +
  geom_segment(aes(x = pred_lower, xend = pred_upper, yend = facility_label),
               color = "grey70", linewidth = 0.8) +
  geom_point(aes(color = occupancy_tier), size = 3) +
  geom_point(aes(x = actual_mean), shape = 4, size = 2, color = "black") +
  scale_color_manual(values = c("High" = "#d7191c", "Medium" = "#fdae61",
                                "Low"  = "#2c7bb6")) +
  scale_x_continuous(labels = scales::comma) +
  theme_minimal() +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(lineheight = 1.3)) +
  labs(
    title    = "Facility Vulnerability Rankings — Predicted 2024 Volume",
    subtitle = "Line = 90% posterior predictive interval; dot = predicted median; x = actual (2024).\nModel underpredicts the highest-volume facilities (actuals fall right of interval).\nColor = baseline occupancy tier (2018-2019). High occupancy + high volume = greatest surge exposure.",
    x = "Predicted Annual Acute Admissions", y = NULL,
    color = "Baseline\nOccupancy"
  )

# Summary table
vulnerability %>%
  arrange(desc(pred_median)) %>%
  mutate(
    across(c(pred_median, pred_lower, pred_upper, actual_mean), ~ round(., 0)),
    residual_mean    = round(residual_mean, 0),
    occupancy_raw    = scales::percent(occupancy_raw, accuracy = 0.1),
    share_60plus_raw = scales::percent(share_60plus_raw, accuracy = 0.1)
  ) %>%
  select(facility_label, pred_median, pred_lower, pred_upper,
         actual_mean, residual_mean, occupancy_raw, share_60plus_raw, occupancy_tier) %>%
  rename(
    Facility         = facility_label,
    `Pred Median`    = pred_median,
    `90% Lower`      = pred_lower,
    `90% Upper`      = pred_upper,
    `Actual Mean`    = actual_mean,
    `Residual`       = residual_mean,
    `Occupancy`      = occupancy_raw,
    `60+ Share`      = share_60plus_raw,
    `Occupancy Tier` = occupancy_tier
  )
```

------------------------------------------------------------------------

# Practical Monitoring Framework

The scorecard model supports a real-time or batch vulnerability monitoring workflow. A facility is flagged as critical when its predicted annual admissions exceed the volume corresponding to 95% bed occupancy, derived from its own licensed bed size and average length of stay (2018-2023).

Capacity threshold per facility: ALOS_j = discharge_days_j / toc_acute_j (avg days per admission) threshold_j = (bed_midpoint_j x 365 x 0.95) / ALOS_j (admissions at 95% occupancy)

```{r monitoring_framework}
# Step 1: compute facility-specific ALOS from full training period (2018-2023)
alos_train <- train %>%
  group_by(oshpd_id, facility_label) %>%
  summarise(
    alos         = mean(dsch_days / toc_acute, na.rm = TRUE),
    bed_midpoint = mean(bed_midpoint,          na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    # Annual admissions at which facility hits 95% licensed bed occupancy
    capacity_threshold = (bed_midpoint * 365 * 0.95) / alos
  )

cat("Facility-specific capacity thresholds (95% occupancy):\n")
alos_train %>%
  mutate(
    alos               = round(alos, 1),
    bed_midpoint       = round(bed_midpoint),
    capacity_threshold = round(capacity_threshold)
  ) %>%
  arrange(desc(capacity_threshold)) %>%
  select(facility_label, bed_midpoint, alos, capacity_threshold) %>%
  rename(
    Facility    = facility_label,
    `Beds`      = bed_midpoint,
    `ALOS`      = alos,
    `Threshold` = capacity_threshold
  ) %>%
  print(n = Inf)

# Step 2: construct forecast data (next year beyond test)
new_data <- test %>%
  filter(data_year == max(data_year)) %>%
  mutate(year_idx = max(year_idx) + 1)

# Step 3: posterior predictive distribution for each facility
pred_new <- posterior_predict(m_scorecard, newdata = new_data)

# Step 4: P(admissions > facility-specific 95% capacity threshold)
surge_risk <- new_data %>%
  left_join(alos_train %>% select(oshpd_id, capacity_threshold),
            by = "oshpd_id") %>%
  mutate(
    pred_median = apply(pred_new, 2, median),
    pred_lower  = apply(pred_new, 2, quantile, 0.05),
    pred_upper  = apply(pred_new, 2, quantile, 0.95),
    p_critical  = map_dbl(seq_len(n()),
                          ~ mean(pred_new[, .x] > capacity_threshold[.x]))
  ) %>%
  arrange(desc(p_critical))

surge_risk %>%
  mutate(
    pred_median        = scales::comma(round(pred_median)),
    capacity_threshold = scales::comma(round(capacity_threshold)),
    p_critical         = scales::percent(p_critical, accuracy = 0.1)
  ) %>%
  rename(
    Facility            = facility_label,
    `Predicted Volume`  = pred_median,
    `95% Cap Threshold` = capacity_threshold,
    `P(Critical)`       = p_critical
  ) %>%
  select(Facility, `Predicted Volume`, `95% Cap Threshold`, `P(Critical)`)
```

```{r monitoring_plot}
surge_risk %>%
  filter(p_critical > 0) %>%
  mutate(
    facility_label = fct_reorder(facility_label, p_critical),
    alert_tier     = case_when(
      p_critical >= 0.50 ~ "Critical (>= 50%)",
      p_critical >= 0.20 ~ "Elevated (20-49%)",
      TRUE               ~ "Watch (< 20%)"
    ),
    alert_tier = factor(alert_tier,
                        levels = c("Critical (>= 50%)", "Elevated (20-49%)", "Watch (< 20%)"))
  ) %>%
  ggplot(aes(x = p_critical, y = facility_label, fill = alert_tier)) +
  geom_col(width = 0.6) +
  geom_vline(xintercept = 0.50, linetype = "dashed", color = "grey30", linewidth = 0.7) +
  geom_text(aes(label = scales::percent(p_critical, accuracy = 0.1)),
            hjust = -0.15, size = 3.2, color = "grey20") +
  scale_x_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_manual(values = c(
    "Critical (>= 50%)"  = "#d7191c",
    "Elevated (20-49%)"  = "#fdae61",
    "Watch (< 20%)"      = "#abd9e9"
  )) +
  theme_minimal() +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(lineheight = 1.3),
        legend.position = "bottom") +
  labs(
    title    = "Surge Risk Monitor — P(Exceeding 95% Capacity) by Facility",
    subtitle = "Probability that forecasted admissions exceed the facility-specific 95% \n bed occupancy threshold. Dashed line = 50% critical threshold. \n Only facilities with P(Critical) > 0 shown.",
    x = "P(Critical)", y = NULL, fill = "Alert Tier"
  )
```
